Computing likelihoods under the SMC for a general class of ARGs.

### 2025-02-20 paper.tex
Stuff

The key contribution of this paper is to rederive a classical
ARG likelihood function in a way that supports the sort of partial information
provided by some modern ARG inference tools,
using a novel backwards-time derivation of the SMC.
The underlying process that we reason about probabilistically is still
composed of events affecting lineages, but, because we
do not necessarily observe or estimate these events,
we integrate over the possible timings implied by the information
in the input ARG.
Although we can associate a likelihood with any ARG in \texttt{tskit} format,
there are some caveats. By treating each edge independently,
our algorithm will still produce a value for ARGs with
the various degrees of completeness to which inheritance
is documented by the different ARG inference methods: e.g., polytomies, absence
of locally unary coalescence nodes, unpreserved identify of internal nodes
across local trees, etc.
However, many of these imprecisions violate the implicit
assumptions of the likelihood computation.
For example, we assume the 
presence of locally unary nodes to correctly identify the
full extent of all lineages going back through time and therefore recombination events.
% Without these unary nodes, our algorithm will detect a recombination
% event along both coalescing lineages in case of a (sub)tree height changes along
% the ARG as both edges would switch parent nodes. 
Locally unary nodes are not a common aspect of current ARG research, but 
are estimated by \texttt{tsinfer}, and can be imputed to high
degree of accuracy from fully simplified ARGs~\citep{fritze2024forest}.

>>>>

The key contribution of this paper is to rederive a classical
ARG likelihood function in a way that supports the sort of partial information
provided by some modern ARG inference tools,
using a novel backwards-time derivation of the SMC.
The underlying process that we reason about probabilistically is still
composed of events affecting lineages, but, because we
do not necessarily observe or estimate these events,
we integrate over the possible timings implied by the information
in the input ARG.
Although we can associate a likelihood with any ARG in \texttt{tskit} format,
there are some caveats. By treating each edge independently,
our algorithm will still produce a value for ARGs with
the various degrees of completeness to which inheritance
is documented by the different ARG inference methods: e.g., polytomies, absence
of locally unary coalescence nodes, unpreserved identify of internal nodes
across local trees, etc.
However, many of these imprecisions violate the implicit
assumptions of the likelihood computation.
For example, we assume the 
presence of locally unary nodes to correctly identify the
full extent of all lineages going back through time and therefore recombination events.
% Without these unary nodes, our algorithm will detect a recombination
% event along both coalescing lineages in case of a (sub)tree height changes along
% the ARG as both edges would switch parent nodes. 
Locally unary nodes are not a common aspect of current ARG 
research~\citep{wong_general_2023}, but 
are estimated by \texttt{tsinfer}, and can be imputed to high
degree of accuracy from fully simplified ARGs~\citep{fritze2024forest}.

Our results, while preliminary, suggest that ``full ARGs``, in which 
each recombination event is precisely pinpointed in time, 
contain a significant degree of redundancy in terms of conveying 
information about the generative process. 
It is possible, therefore, that explicitly estimating these 
events is not necessary for either inference under the SMC or 
for downstream applications, and we can integrate over the 
plausible times as we do here.
Further research is required to fully understand the
